Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-19 inference.tex
Explanation added.

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation $\in$ to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
($\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similar.

The first-order encoding of a bind expression describes the semantics
of set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
translation first encodes $s_1$ and $s_2$ to obtain predicates
$(\phi_1,\G_1)$ and $(\phi_2,\G_2)$, respectively. Since $s_2$ is
under a new binder that binds $x$, free variable sequence of $s_2$ is
$\nubar,x$. For technical reasons, we substitute a fresh $a$ for $x$,
making the sequence $\nubar,a$. The set $s$ is encoded as a a new
uninterpreted predicate $g$ indexed by $s$'s free variables
($\nubar$). Since the set denoted by $g$ is the result of the bind
$s_1 \bind \lambda x.s_2$, a first-order constraint defining the bind
operation (as described above) is generated. The constraint relates the
predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to
the uninterpreted predicate $g$ that represents $s$.

>>>>

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation $\in$ to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
($\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similarly.

The first-order encoding of a bind expression describes the semantics
of set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
translation first encodes $s_1$ and $s_2$ to obtain predicates
$(\phi_1,\G_1)$ and $(\phi_2,\G_2)$, respectively. Since $s_2$ is
under a new binder that binds $x$, free variable sequence of $s_2$ is
$\nubar,x$. For technical reasons, we substitute a fresh $a$ for $x$,
making the sequence $\nubar,a$. The set $s$ is encoded as a a new
uninterpreted predicate $g$ indexed by $s$'s free variables
($\nubar$). Since the set denoted by $g$ is the result of the bind
$s_1 \bind \lambda x.s_2$, a first-order constraint defining the bind
operation (as described above) is generated. The constraint relates
the predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$
(resp.), to the uninterpreted predicate $g$ that represents $s$. The
constraint is assigned a name ($\pi$) so that it can be enabled only
for the choice of free variables ($\nubar$) for which $g$ needs to be
defined; observe that the lambda returned by
$\mssemof{\nubar}{s_1 \bind \lambda x.s_2}$ asserts $\pi(\vbar)$ only
in conjunction with $g(\vbar,r)$, for some $r$.

The first-order encoding of the $\existsl$ expression eliminates the
existential by Skolemizing it. Skolemizing is the process of
substituting an existentially bound $x$ in $\phi\in\Prop^1$ with
$f(\nubar)$, where $f$ is a fresh uninterpreted function, and $\nubar$
are the free variables in $\phi$ bound by enclosing universal
quantifiers. Due to the decidability restrictions that will become
clear shortly, the only uninterpreted functions we admit in our logic
are boolean (i.e., predicates/relations). Consequently, Skolemizing an
existentially bound $x:T$ with $f(\nubar):T$, where $T$ is a
non-boolean type, is not really an option. Instead, we define $f$ as
an uninterpreted relation that behaves as a partial function:
\begin{smathpar}
  \begin{array}{c}
    \forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
    \Rightarrow a = b
  \end{array}
\end{smathpar}
We then replace the existentially bound $x$ with a new universally
bound $a$ in $\phi$, under the condition that $f(\nubar,a)$ holds.
Due to the partial function property of $f$, there can be utmost one
instantiation for $a$, as desired. There can however be no
instantiations, i.e., no $a$ that satisfies $f(\nubar,a)$, thus
violating the exists property. Fortunately, this problem can be solved
in our case by observing that the existential always occurs in the
context (antecedent) in the verification conditions (implications)
generated by the \rulelabel{RG-Txn} rule. Intuitively, we prove that a
set of records is equal to, or contained in, other set of records
(e.g., $\F_1 = \F_2$ or $\F \subseteq G$) \emph{assuming} that there
exists a state $\stg$ that satisfies $I$. If there does not exist such
a state, then our premise itself is false, and the goal must be
vacuously provable. We make use of this observation, and let the
context derive $\bot$ when there does not exist a desired $a$.
